Higher order unification for free!

### 2024-04-08 paper/main.tex
wip tex: Add decompilation

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial, since the terms are beta normal beta just builds an app.
\end{proof}

@@

% Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
% terms and, consequently replace them with fresh \Ho unification variables at
% compilation time. The code below illustrate how this relation is used to for
% term compilation.
The following rule is inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}.

@@

\noteD[inline]{Dire della commit-links: intuizione fa round trip}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is performed between
  terms in \wellb, therefore, the substitution remains \wellb.
\end{proof}

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
  terms. This is equivalent to \elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
\end{invariant}

@@

Detection of \maybebeta is quite simple to implement in the compiler, since it
is sufficient to detect applications with flexible head and argument that
are not in \llambda.

@@

\noteD[inline]{Dire della commit-links: intuizione non devono esserci beta per corollary}

@@

\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}}
\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}}

@@

@@

\noteD[inline]{Dire qualcosa sulla commit-links}

>>>>

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial, since the terms are beta normal beta just builds an app. \noteD{Reformulate?}
\end{proof}

@@

% Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
% terms and, consequently replace them with fresh \Ho unification variables at
% compilation time. The code below illustrate how this relation is used to for
% term compilation.

\paragraph{Compilation}
The following rule is inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}.

@@

\paragraph{Decompilation}
Decompilation of \linketa is performed by adding new rules to the
\elpiIn{commit-link} predicate. In particular, given \linketaM{\Gamma}{X}{t}, we
can note that this unification never fails, since $X$ is a flexible term and no
other \linketa has $X$ has \lhs (by \cref{def:progressetadedup}). The link is
remove from \linkStore and \elpiIn{commit-links} terminates.

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-2,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is performed between
  terms in \wellb, therefore, the substitution remains \wellb.
  \noteD[inline]{Bisogna aggiungere un lemma nella \cref{sec:nutshell} che dice che
  unificare due termini in \wellb, in una $\sigma$, tale che $\wellb(\sigma)$, 
  non invalida \wellb}
\end{proof}

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$ and $t_1\dots t_m$ is a list of
  terms. This is equivalent to \elpiIn{app[uva X S | L]}, where \elpiIn{S}~$=
  s_1\dots s_n$ and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

\paragraph{Compilation}
Detection of \maybebeta is quite simple to implement in the compiler, since it
is sufficient to detect applications with flexible head and argument that
are not in \llambda.

@@

\paragraph{Decompilation}
During \elpiIn{progress}, as claimed in \cref{inv:beta-rhs}, the decompilation
can only have \linkbeta with not instantiated \lhs. In this case, \lhs is
unified with \rhs.
\noteD[inline]{not really sure of this, we can have $F\appsep a = \lambda x.G x$.
In this case when do we fail: for sure in decompile. But to respect fidelity,
we should fail immediately: we have a \linkbeta and a \linketa with same \lhs}

@@

\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}}
\newcommand{\progBetaRH}{\emph{progress-beta-rigid-head}}
\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}}

@@

\noteD{added the def below to respect fidelity}
% Example of this `X a = Y a'
\begin{definition}[\progBetaDedup]
  Given two \linkbeta $l1$ and $l2$, if they share the same \lhs, unification
  fails.
  \label{def:progBetaDedup}
\end{definition}

@@

\noteD{Paragraph for decompile in case of \maybebeta}
The \elpiIn{commit-link} predicate can be extended to add heuristics if during
the decompilation phase \linkbeta remain. For example, the same approximation
explained above can be delayed and applied only if the terms in \maybebeta never
falls in \llambda after the execution of all the unification problems. We want
to point out, that we call this \emph{approximation}, since we are
making a choice among all the possible unifiers and therefore, we can pick
the wrong one.
